---
output:
  html_fragment:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

I like visual arts, but I'm moderately colorblind and thus have never been great at making my own works. When I'm plotting data and need colors, my standard procedure is having a website generate me a color palette, or finding a visually pleasing one someone else has made and posted online.  

I also love film, and I started thinking about ways I could generate color palettes from films that use color beautifully. There are a [number of packages](https://towardsdatascience.com/generating-color-palettes-in-r-64394117a662) that can generate color palettes from images in R, but I wanted to try writing the code myself.  

I also wanted to not just *generate* a color palette from an image, but then *swapping* it with a different color palette from a different film. This is similar to [neural style transfer with TensorFlow](https://www.tensorflow.org/tutorials/generative/style_transfer), but much simpler. I'm one of those people that likes to joke how [OLS is undefeated](https://twitter.com/mjbsp/status/1521930611052666882); I generally [praise the use of simpler models](https://twitter.com/markhw_/status/1521455082881048576) over more complex ones. So instead of a neural network, I use k-means clustering to transfer a color palette of one still frame from a film onto another frame from a different movie.  

## Functions

Here's the code for the functions I'll be using. I'll describe them in more detail below.  

```{r functions, eval=FALSE}
library(imager)
library(tidyverse)

norm <- function(x) (x - min(x)) / (max(x) - min(x))

shuffle <- function(x) x[sample(seq_along(x), length(x))]

get_palette <- function(filename, k, mdn = FALSE) {
  
  dat_pal <- load.image(filename) %>% 
    as.data.frame(wide = "c")
  
  res_pal <- dat_pal %>% 
    select(starts_with("c")) %>% 
    kmeans(k, algorithm = "Lloyd", iter.max = 500)
  
  if (!mdn) {
    pal <- res_pal$centers %>% 
      as_tibble() %>% 
      mutate(hex = rgb(c.1, c.2, c.3)) %>% 
      pull(hex)
  } else if (mdn) {
    pal <- dat_pal %>% 
      mutate(cluster = res_pal$cluster) %>% 
      group_by(cluster) %>% 
      summarise(across(starts_with("c"), median)) %>% 
      mutate(hex = rgb(c.1, c.2, c.3)) %>% 
      pull(hex)
  }
  
  return(pal)
}

make_plot <- function(filename_in, pal, xy = TRUE) {
  
  the_shot <- load.image(filename_in)
  
  dat_shot <- the_shot %>% 
    as.data.frame(wide = "c")
  
  dat_shot_norm <- dat_shot %>% 
    when(!xy ~ select(., starts_with("c")), ~ .) %>% 
    mutate(across(everything(), norm))
  
  res_shot <- kmeans(
    dat_shot_norm, 
    length(pal), 
    algorithm = "Lloyd", 
    iter.max = 500
  )
  
  dat_shot$clust <- factor(res_shot$cluster)
  
  p <- ggplot(dat_shot, aes(x = x, y = y)) +
    geom_raster(aes(fill = clust)) +
    scale_y_reverse() +
    theme_void() +
    theme(legend.position = "none") +
    scale_fill_manual(values = pal)
  
  return(list(plot = p, dims = dim(the_shot)[1:2]))
}
```

When I thought about transferring the color of one film onto an image from another film, two things came to mind immediately. *The Umbrellas of Cherbourg* is one of the most visually striking films I've ever watched; there's such a dazzling variety of colors, and it displays a wide collection of unique wallpaper. As for what shot to impose those colors onto, one of my favorite shots is the "coffee scene" from [*Chungking Express*](https://www.youtube.com/watch?v=xLpnleQbsHY).  

The function `get_palette()` reads an image in using the `{imager}` package. This package allows you to decompose the image into a data frame, where each row is a pixel. There are `x` and `y` columns, which show where at in the image the pixel is when plotted. There are three additional columns that contain the RGB values. I k-means cluster the three RGB columns, using the built-in `kmeans` function and giving it an arbitrary `k`, and extract the average RGB values from each cluster (i.e., the cluster centers) and then convert them to hex values using the built-in `rgb` function.  

And then `make_plot()` works similarly. It takes a file name, reads it in, and converts it to a data frame. This time, I allow the `x` and `y` columns to be used in the clustering. This means that clustering will be a mix of (a) what the original color was, and (b) where at in the frame the pixel is. All columns are normalized. I use the length of the color palette to determine `k`. I then plot it with `{ggplot2}`, using the new color palette to fill-in according to the clustering of the new pixels. It's more or less a coloring book, where the lines are determined by k-means clustering.    

```{r plot1, eval=FALSE}
set.seed(1839)

pal1 <- get_palette("umbrellas.jpeg", k = 12)
plot1 <- make_plot("chungking.jpeg", pal = pal1)

ggsave(
  "chungking_k12.png", 
  plot1$plot, 
  width = plot1$dims[1], 
  height = plot1$dims[2],
  units = "px"
)
```

I write the file out to the same dimensions to preserve the integrity of the aspect ratio. Here's the two original shots, and then the one produced with `make_plot()`:  

[INSERT UMBRELLAS]

[INSERT CHUNGKING]

[INSERT BLEND]

We can see that the new coloring is a blend of pixel location and the color of the original pixel.  

I started playing around with other ideas, and include two new parts in this next image blend. First, I get the median value, instead of the mean, of the RGB values when clustering for the palette; and second, I shuffle up the order of the palette randomly before feeding it into the plotting function.  

I wanted to apply a movie with warm colors to a movie with cool colors. My mind went to *Her* and *Blade Runner*, respectively.  

```{r plot2, eval=FALSE}
set.seed(1839)

pal2 <- get_palette("her.jpeg", k = 3, mdn = TRUE)
plot2 <- make_plot("bladerunner.jpeg", pal = shuffle(pal2))

ggsave(
  "bladerunner_k3.png", 
  plot2$plot, 
  width = plot2$dims[1],
  height = plot2$dims[2],
  units = "px"
)
```

The originals:

[INSERT HER]

[INSERT BLADE RUNNER]

And the blend:

[INSERT BLEND]

What I like about this is that, since we include `x` and `y` in the clustering of the second image, we get different colors on either side of Roy Batty's face.  

I also wanted to see what the influence of taking out the `x` and `y` values would be. `xy = FALSE` removes any influence of where the pixel is placed in the image, so clustering is done purely on RGB values.  

```{r plot3, eval=FALSE}
set.seed(1839)

pal3 <- get_palette("2001.jpeg", k = 5, mdn = TRUE)
plot3 <- make_plot("arrival.jpeg", pal = shuffle(pal3), xy = FALSE)

ggsave(
  "arrival_k5.png", 
  plot3$plot, 
  width = plot3$dims[1],
  height = plot3$dims[2],
  units = "px"
)
```

I wanted to combine these two shots from *2001: A Space Odyssey* and *Arrival* because they visually rhyme with one another:  

[INSERT 2001]

[INSERT ARRIVAL]

We can see in the color-blended image that colors fill in on spaces that are separated geographically from one another in the xy-plane of the image:  

[INSERT BLEND]

Compare this to another version I made, where I allowed `x` and `y` to be included:  

[INSERT OTHER BLEND]

We see that vertical line in the upper third of the shot forming due to the influence of `x` in the data. This also demonstrates overfitting: It'll draw a line where two adjacent data points are functionally equivalent if you misspecify k. But for aesthetic purposes, overfitting isn't necessarily a problem!  

We also see a indistinct boundaries of one color into another here. The underlying image has few distinct lines—the entire image is ink drawn onto a wispy mist. What about when we get distinct lines and contrast? The easy answer for clean lines would have been to go to Wes Anderson here, but I felt like that was too expected from a blog post written by somebody such as myself. So instead, I took colors from the animated *Lion King*, a vibrant film, and projected it onto one of Roger Deakins' best shots from *Fargo*.  

```{r plot4, eval=FALSE}
set.seed(1839)

pal4 <- get_palette("lionking.jpeg", k = 2, mdn = TRUE)
plot4 <- make_plot("fargo.jpeg", pal = rev(pal4), xy = FALSE)

ggsave(
  "fargo_k2.png", 
  plot4$plot, 
  width = plot4$dims[1],
  height = plot4$dims[2],
  units = "px"
)
```

[INSERT LION KING]

[INSERT FARGO]

[INSERT BLEND]

The last thing I wanted to do was look at a shot that had two primary colors and project it onto a black-and-white film, replacing that underlying dichotomy with two other colors.  

```{r plot5, eval=FALSE}
set.seed(1839)

pal5 <- get_palette("killbill.jpeg", k = 2, mdn = TRUE)
plot5 <- make_plot("strangelove.jpeg", pal = pal5, xy = FALSE)

ggsave(
  "strangelove_k2.png", 
  plot5$plot, 
  width = plot5$dims[1],
  height = plot5$dims[2],
  units = "px"
)
```

This shot from *Kill Bill Vol. 1* came to mind for a shot that was mostly two colors:  

[INSERT KILL BILL]

While I went with my favorite scene from *Dr. Strangelove*, perhaps the funniest film ever made, for the black-and-white still:  

[INSERT STRANGELOVE]

[INSERT BLEND]

The functions are above and the full code is at [my GitHub](https://github.com/markhwhiteii/blog/tree/master/filmstills). Try playing with the functions and blending images; it's fun, but it also a visual guide that helps you truly understand what exactly k-means clustering is doing.  
