---
output: html_fragment
---

```{r setup, include = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(knitr)
prettify <- function(x) {
  sapply(unlist(strsplit(x, "_")), function(y) {
    substr(y, 1, 1) <- toupper(substr(y, 1, 1))
    y
  }) %>% 
    paste(collapse = " ")
}
set.seed(1839)
n <- 100000
x <- runif(n, .501, .999)
y <- rbinom(n, 1, x)
simdat <- data.frame(x, y)

fig_simulated <- ggplot(simdat, aes(x = x, y = y)) +
  geom_smooth(method = "lm", se = FALSE, color = "purple") +
  geom_smooth(method.args = list(family = binomial()), 
              se = FALSE, color = "gold", method = "glm") +
  theme_minimal()

dat <- read_csv("fivethirtyeight_final.csv") %>% 
  filter(win_probability < 1L)

results <- with(
  dat[!is.na(dat$won), ],
  list(
    mean_win_prob = mean(win_probability),
    correct_rate = mean(won)
  )
) %>% 
  lapply(function(x) paste0(round(x, 3) * 100, "%"))

outstanding_races <- dat %>% 
  filter(is.na(won)) %>% 
  select(state, race_type, candidate, party, incumbent, win_probability)

wrong_predictions <- dat %>% 
  filter(!won & !is.na(won)) %>% 
  arrange(desc(win_probability)) %>% 
  select(state, race_type, candidate, party, incumbent, win_probability)

figure1 <- ggplot(dat, aes(x = win_probability, y = as.numeric(won))) +
  geom_point() +
  geom_smooth(
    method = "glm", 
    method.args = list(family = binomial), 
    se = FALSE,
    color = "gold"
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "purple"
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  labs(x = "538 Predicted Win Probability", y = "Actual Win Probability") +
  theme_minimal() +
  theme(text = element_text(size = 16))

# what if all outstanding races are wrong?
figure2 <- dat %>% 
  mutate(won = ifelse(is.na(won), FALSE, won)) %>% 
  ggplot(aes(x = win_probability, y = as.numeric(won))) +
  geom_point() +
  geom_smooth(
    method = "glm", 
    method.args = list(family = binomial),
    se = FALSE,
    color = "gold"
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "purple"
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  labs(x = "538 Predicted Win Probability", y = "Actual Win Probability") +
  theme_minimal() +
  theme(text = element_text(size = 16))

# bucketed
results_thresholds <- dat %>% 
  filter(win_probability <= .75 & !is.na(won)) %>% 
  summarise(threshold = .75,
            count = n(),
            mean_win_probability = mean(win_probability),
            correct_rate = mean(won)) %>% 
  bind_rows({
    dat %>% 
      filter(win_probability <= .65 & !is.na(won)) %>% 
      summarise(threshold = .65,
                count = n(),
                mean_win_probability = mean(win_probability),
                correct_rate = mean(won))
  }) %>% 
  bind_rows({
    dat %>% 
      filter(win_probability <= .55 & !is.na(won)) %>% 
      summarise(threshold = .55,
                count = n(),
                mean_win_probability = mean(win_probability),
                correct_rate = mean(won))
  })
```

The average 538 predicted probability of a correct prediction was `r results$mean_win_prob`, and the average actual correct prediction rate was `r results$correct_rate`.  

The outstanding races, as of publishing, are:  

<div align="center">
```{r outstanding, echo = FALSE}
kable(
  outstanding_races %>% 
    mutate_if(is.character, function(x) lapply(x, prettify)),
  col.names = sapply(names(outstanding_races), prettify),
  digits = 3,
  format = "html"
)
```
</div>

All the incorrect 538 predictions are:  

<div align="center">
```{r wrong_predictions, echo = FALSE}
kable(
  wrong_predictions %>% 
    mutate_if(is.character, function(x) lapply(x, prettify)),
  col.names = sapply(names(wrong_predictions), prettify),
  digits = 3,
  format = "html"
)
```
</div>

If the 538 model was completely accurate in its quantification of win probability, we would expect the relationship between estimated win probability (x-axis) and actual win probability (y-axis) to look like:  

```{r simulated, echo = FALSE, out.width = '70%', fig.align = 'center'}
fig_simulated
```

And what it actually looked like:  

```{r fig1, warning =  FALSE, echo = FALSE, out.width = '70%', fig.align = 'center'}
figure1
```

If all of the outstanding predictions were *wrong* (i.e., the worst-case scenario for 538), the relationship would be:  

```{r fig2, echo = FALSE, out.width = '70%', fig.align = 'center'}
figure2
```

<div align="center">
```{r threshold_results, echo = FALSE, results = "asis"}
kable(
  results_thresholds %>% 
    mutate_if(is.character, function(x) lapply(x, prettify)),
  col.names = sapply(names(results_thresholds), prettify),
  digits = 3,
  format = "html"
)
```
</div>
